\vspace{-3em} \textbf{ Mapping stellar surfaces\\ II: An interpretable Gaussian process model for light curves }

### 2021-02-01 paper2/ms.tex
chomments

In this section, we provide a brief overview of Gaussian processes
(\S\ref{sec:gp-intro}) and spherical harmonics (\S\ref{sec:gp-ylms}),
followed by an outline of the derivation of our interpretable GP
(\S\ref{sec:gp-gp}).
This derivation boils down to computing the mean and covariance of the stellar
flux conditioned on certain physical properties of the star and its starspot
distribution.
In \S\ref{sec:gp-inc} and \S\ref{sec:gp-norm} we derive useful extensions
of the model. For convenience, we summarize the results of this entire section in
\S\ref{sec:gp-summary}.
Most of the math is folded into the Appendix for readability;
readers may want to refer to Appendix~\ref{sec:notation} in particular for
a discussion of the notation and conventions we adopt.

@@

@@

(Appendix \ref{sec:lat}). Finally,

@@

Figure~\ref{fig:calibration_default_latitude} shows samples from the spot
latitude posterior (hyper)distribution. Since the parameters $\mu_\phi$
and $\sigma_\phi$ characterize a distribution over spot latitudes, uncertainty
in their values translates to uncertainty in the actual shape of the spot
latitude distribution. Thus, the ensemble of blue curves in
Figure~\ref{fig:calibration_default_latitude} quantifies our knowledge of
how spots are distributed on the surfaces of the stars in the dataset.
These distributions are again consistent with the true distribution used
to generate the spots (orange curve) within less than 2 standard deviations.

@@

The runs so far correspond to stars with many ($n = 20$) spots, for which the
resulting light curves are smooth due to the fact that many spots are in
view at any given time. Figures~\ref{fig:calibration_hicontrast} and
\ref{fig:calibration_bigspots} show what happens when the model is applied to
stars with $n=2$ and $n=1$ spots, respectively. Despite large portions of the
light curves being flat (and therefore extremely non-stationary)
in these scenarios, the GP does surprisingly well, recovering the radii
and latitude parameters within $2-3\sigma$ in both cases.
Note that in order to preserve the same signal-to-noise ratio relative to
the other runs, we gave the spots in Figure~\ref{fig:calibration_hicontrast}
a much higher contrast ($c = 0.5$). Even though the contrast is degenerate
with the number of spots (which is very poorly constrained), the $c$ posterior
has a much heavier tail than in the other runs.
Thus, in spite of the arguments in \citet{PaperI} about the difficulty in
constraining $c$ and $n$ from single-band photometry, it is evident that
the full covariance structure of the data encodes \emph{some} information
about the contrast and--to a much lesser extent--the number of spots.
In Figure~\ref{fig:calibration_bigspots}, we compensate for the smaller number
of spots by increasing the spot radius to $r = 45^\circ \pm 5^\circ$ instead,
showing that the model can accurately model large spots, even in the
presence of some (unmodelled) scatter in their sizes.

@@

We therefore conclude that \textbf{relative single-band photometry is largely insensitive to
    the difference between bright spots and dark spots}. However, it is
important to bear in mind that the
degeneracy described above exists only for the \emph{Gaussian approximation}
to the likelihood function. As we argued earlier, the true likelihood function
is not a Gaussian; in particular, the true probability distribution
has higher-order moments that we do not model here. These moments should
in principle encode information about the sign of the spot contrast, but they
may be very difficult to infer in practice. It may be possible to distinguish
between dark spots and bright spots with traditional forward models of stellar
surfaces, but (as we argued earlier) a statistically rigorous ensemble analysis
of stellar light curves using such forward models is probably computationally
intractable.

>>>>

In this section, we provide a brief overview of Gaussian processes
(\S\ref{sec:gp-intro}) and spherical harmonics (\S\ref{sec:gp-ylms}),
followed by an outline of the derivation of our interpretable GP
(\S\ref{sec:gp-gp}).
This derivation boils down to computing the mean and covariance of the stellar
flux conditioned on certain physical properties of the star and its starspot
distribution.
In \S\ref{sec:gp-inc} and \S\ref{sec:gp-norm} we derive useful extensions
of the model. \textbf{For convenience, we summarize the results of this entire section in
    \S\ref{sec:gp-summary}.}
Most of the math is folded into the Appendix for readability;
readers may want to refer to Appendix~\ref{sec:notation} in particular for
a discussion of the notation and conventions we adopt.

@@

Finally,

@@

(Appendix \ref{sec:lat}). This allows us to model starspot distributions
with ``active latitudes'' of tunable width that are symmetric about
the equator. The distribution is flexible enough to also model equatorial
spots and isotropically-distributed spots. Stars with multiple active
latitudes can easily be modeled as a sum of Gaussian processes
(\S\ref{sec:mixture}). These choices for the spatial distribution of
spots are based on the Sun, whose spots emerge in azimuthally-symmetric
belts at roughly the same latitude in both hemispheres, then
migrate toward the equator over the course of the 11-year cycle
\citep{Solanki2006}.

@@

Figure~\ref{fig:calibration_default_latitude} shows samples from the spot
latitude posterior (hyper)distribution. Since the parameters $\mu_\phi$
and $\sigma_\phi$ characterize a distribution over spot latitudes, uncertainty
in their values translates to uncertainty in the actual shape of the spot
latitude distribution. Thus, the ensemble of blue curves in
Figure~\ref{fig:calibration_default_latitude} quantifies our knowledge of
how spots are distributed on the surfaces of the stars in the dataset.
These distributions are again consistent with the true distribution used
to generate the spots (orange curve) within less than 2 standard deviations.%
\footnote{%
    If the results in Figure~\ref{fig:calibration_default_latitude}
    seem biased, recall from Figure~\ref{fig:calibration_default_corner}
    that the mean of the latitude distribution is
    consistent with the truth at $2-3\sigma$. That is roughly the difference
    between the orange curve and the average of the blue curves. As we
    will see, inference with a larger ensemble
    (Figure~\ref{fig:calibration_default_1000})
    allows us to infer the mean latitude to within about two degrees.
}

@@

The runs so far correspond to stars with many ($n = 20$) spots, for which the
resulting light curves are smooth due to the fact that many spots are in
view at any given time. Figures~\ref{fig:calibration_hicontrast} and
\ref{fig:calibration_bigspots} show what happens when the model is applied to
stars with $n=2$ and $n=1$ spots, respectively. Despite large portions of the
light curves being flat (and therefore extremely non-stationary)
in these scenarios, the GP does surprisingly well, recovering the radii
and latitude parameters within $2-3\sigma$ in both cases.
Note that in order to preserve the same signal-to-noise ratio relative to
the other runs, we gave the spots in Figure~\ref{fig:calibration_hicontrast}
a much higher contrast ($c = 0.5$). Even though the contrast is degenerate
with the number of spots (which is very poorly constrained), the $c$ posterior
has a much heavier tail than in the other runs.
Thus, in spite of the arguments in \citet{PaperI} about the difficulty in
constraining $c$ and $n$ from single-band photometry, it is evident that
the full covariance structure of the data encodes \emph{some} information
about the contrast and--to a much lesser extent--the number of spots.
In Figure~\ref{fig:calibration_bigspots}, we compensate for the smaller number
of spots by increasing the spot radius to $r = 45^\circ \pm 5^\circ$ instead,
showing that the model can accurately model large spots, even in the
presence of some (unmodeled) scatter in their sizes.

@@

We therefore conclude that \textbf{single-band photometry is largely insensitive to
    the difference between bright spots and dark spots}. However, it is
important to bear in mind that the
degeneracy described above exists only for the \emph{Gaussian approximation}
to the likelihood function. As we argued earlier, the true likelihood function
is not a Gaussian; in particular, the true probability distribution
has higher-order moments that we do not model here. These moments should
in principle encode information about the sign of the spot contrast, but they
may be very difficult to infer in practice. It may be possible to distinguish
between dark spots and bright spots with traditional forward models of stellar
surfaces, but (as we argued earlier) a statistically rigorous ensemble analysis
of stellar light curves using such forward models is probably computationally
intractable.
